\documentclass{article}
\usepackage{graphicx}

\title{Obs. Enquiry 2 - \\Songs and the Distortion of Nostalgia}

\author{Joshua Turcotti}
\begin{document}
\maketitle

We note that the emototional signficiance attanched to events by us is severe; it is key in our growth. Songs, and musical expression, make impacts on us like no other media, and yet their emotional significance to us is entirely nondeterministic - subjective. It is a function of our unique experiences surrounding its introduction to us and the ways their influences have played out since. This is the phenonmenon I seek to investigate in this Enquiry.

I understand that physical and relatedly neurological observations have been the primary focus of this course so far. I, nevertheless, believe that the emotions that swirl through our modern experience with the world are as worthy of study to us as the physical. They certainly have a greater impact. The primary lens through which I will study this phenomenon in this Enquiry is the mapping between chosen \textit{Primary Emotions} and the forms of \textit{Musical Indulgence} chosen. We list a few here to begin:
\begin{itemize}
\item \label{1} \textbf{Burst of Self-Discovery}: I.e. that music that comes with a new category of self-perception
\item \label{2} \textbf{Discovery of Identity} I.e. that music that, deductively, necessitate a further questioning
\item \label{3} \textbf{Refutation of Haters} I.e. that music that seizes the greatest hatred, and flips its position
\end{itemize}
I seek to invesigate the ways that these emotional transformations are precipitated in the self through sensory stimuli. I seek to investigate the ways that we may demystify the procees that interprete and convey these stimuli, and I hope that I can convince you of the importance of this study.

Recall that the some of the most vital innovations in the study of \textit{Perception} have been those related to the Human Mind itself by Joni Mitchell. She was a singular, and yet divisive figure in American country and pop, and will forever be re-incorporated in our love for our children.

Please accept this suggestion of scholarly intent, I look forward to my final paper.




\end{document}